---
title: "Plotting SST with ggplot"
author: Eric C. Anderson
date: "2014-12-24"
slug: test-post
categories: []
tags: []
output:
  default:
    toc: true
---



Roy M and I were curious about how one might go about efficiently and quickly
plotting big rasters like sea surface temperature data using `ggplot` with nice
coastlines in there, etc.

I had considered using `ggmap`.  For reasons to do with the mercator projection of
the base map in `ggmap` the word on the street is that this doesn't work very
well.  I regularly crashed my whole R session trying it, too!  So, instead
we will focus on using the [GSHHS](http://www.ngdc.noaa.gov/mgg/shorelines/gshhs.html) shoreline maps.


## A Catalina Island Example

I am going to be heading out to Catalina Island for some snorkeling with the
family over Xmas break.  A lot of friends have told me that the water there
is warmer than typical this time of year.  It seems this would make a good
test case. So, here is my plan:

1. First figure out how to plot this year's 8-day composite SST for a large
area around Catalina Island, using `ggplot2`.
2. Then, I want to compare across, say, 9 years, using faceting in ggplot.

Before we get going, here are the libraries that I will be using:
```{r}
library(stringr)
library(ncdf4)
library(reshape2)
library(ggplot2)
library(mapdata)
library(maptools)
```


### Getting SST data

Roy showed me how to download the data.  I am going to wrap that up in a little
function that downloads data from Dec 12 in the year that I tell it to, and saves
that in an intermediate file, so it doesn't have to re-download it. Note that,
since the other things we will be plotting assume that western hemisphere longitudes
are negative, we subtract 360 from the SST longitudes. I also add the date to the
output list, as I will want to use that later.
```{r}
# This is based on code that Roy sent me for getting SST.
# I modified it to get data from around Catalina Island
my_get_sst_catalina <- function(year) {
  
  year <- as.character(year)
  # base url with YEAR meant to be replaced
  turl <- "http://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMWsstd8day.nc?sst[(YEAR-12-12T00:00:00Z)][(0.0)][(30.0):(35.0)][(238.0):(243.2)]"
  
  # the URL with YEAR replaced with a value
  the_url <- str_replace(turl, "YEAR", year)
  
  # the filename to save the downloaded data in.  I put this in a cache folder
  # a few directory levels above...
  the_file <- paste("../../file_caches_etc/sst_", year, ".nc", sep = "")
  
  # if the file isn't here, download it
  if (!file.exists(the_file)) {
    download.file(the_url, the_file, mode = "wb")
  } else {
    message(paste("Using existing file", the_file, collapse = " "))
  }
  
  # now, grab stuff out of the netcdf file and return it in a list
  # called ret
  sstFile <- nc_open(the_file)
  ret <- list()
  ret$lats <- ncvar_get(sstFile, "latitude")
  ret$lon <- ncvar_get(sstFile, "longitude") - 360 # we need them as negative values
  ret$time <- ncvar_get(sstFile, "time")
  ret$sst <- ncvar_get(sstFile, "sst")
  ret$date <- paste("12-12-", year, sep = "")
  nc_close(sstFile)
  
  ret
}
getwd()
```

Here is an example of how to use that function, and a summary of the output.
```{r, message=FALSE}
cata_sst_2014 <- my_get_sst_catalina(2014)

# look at the structure of the result
str(cata_sst_2014)
```
The sst component is a matrix with elements that correspond to cells of latitude
and longitude.  It is a raster.

### Prepping for ggplot and a simple plot

You might recall that ggplot operates on long-format data frames.  A raster
in matrix format is in wide format.  So, we need to convert that. We can
do that pretty easily with the `reshape2` package.  The trick here is
to put the long's and lat's on as row and column names and then melt it. I will
wrap this up in a function so we can use it easily later.  I will also
cbind the date to the result, and make a column for Fahrenheit because when
most people (in the backwards, non-metric US) are thinking of
swimming in cold water, they gauge temp in Fahrenheit (myself included...):
```{r}
melt_sst <- function(L) {
  dimnames(L$sst) <- list(long = L$lon, lat = L$lats)
  ret <- melt(L$sst, value.name = "sst")
  cbind(date = L$date, ret, degF = ret$sst * 9/5 + 32)
}
```
And here is how we use that function.  And a look at what it returns.
```{r}
msst <-  melt_sst(cata_sst_2014)
head(msst)
```

Now that we have a long format data frame, we can plot the contents of it
with `geom_raster` in ggplot.  (Note that most people think of raster data as
being in a matrix format, but, like all things ggplot, `geom_raster` still requires
the data to be in long format). 

Here is a simple plot:
```{r super-vanilla-sst}
ggplot(data = msst, aes(x = long, y = lat, fill = degF)) + 
  geom_raster(interpolate = TRUE) +
  scale_fill_gradientn(colours = rev(rainbow(7)), na.value = NA) +
  theme_bw() +
  coord_fixed(1.3)
```

That is standard ggplot stuff.  Worth mentioning though is `coord_fixed` which
sets the aspect ratio to something that looks good in this part of the world, and
the `na.value` argument to `scale_fill_gradientn` which ensure that missing data
areas don't get any color on them.


### Adding some coastline

The plot above is sort of nice, but it is ugly to have the coast being the same non-color as missing data, and there is not clean edge on the coast.  We will fix that.

A previous post (need to link) showed how to use the [GSHHS](http://www.ngdc.noaa.gov/mgg/shorelines/gshhs.html) shoreline maps.  We use it here too.

Here we get the data that we need for that.  Note that if you want to do this
yourself, you would need to download the maps and probably change the path
to them below:
```{r, message=FALSE, error=FALSE, warning=FALSE}
if (!rgeosStatus()) gpclibPermit()

# path to the GSHHS maps on my computer
gshhs.f.b <- "/Users/eriq/Maps/gshhg-bin-2.3.3/gshhs_f.b"
shore <- getRgshhsMap(gshhs.f.b, xlim = c(-122, -116.8), ylim = c(30, 35))
shore <- fortify(shore)
```
The `fortify()` command turns the polygons from the GSHHS maps into a long
data format that ggplot can deal with.  Here is what `shore` looks like now:
```{r}
head(shore)
```

Now, adding that coastline in there is as simple as just laying down the 
polygons in another layer.  We give it a gray fill and a black coastline.
```{r basic-plot-with-shore}
basic_plot <- ggplot(data = msst, aes(x = long, y = lat, fill = degF)) + 
  geom_raster(interpolate = TRUE) +
  geom_polygon(data = shore, aes(x = long, y = lat, group = group), color = "black", fill = "grey80") +
  scale_fill_gradientn(colours = rev(rainbow(7)), na.value = NA) +
  theme_bw() +
  coord_fixed(1.3)

basic_plot
```



### Harnessing the power of ggplot

The above is pretty, but recall that I want to compare temperatures on Dec 12
this year, to temps in the 8 previous years.  One of the lovely things about
using ggplot to plot our sea surface temperatures is that we have immediate access 
to all of ggplot's features like faceting.

So, let's do this.  First we have to download all the data for 2006-2014 and
put it all together into a single data frame.  Like so:
```{r, message=FALSE}
years <- 2006:2014
tmp <- lapply(years, function(x) melt_sst(my_get_sst_catalina(x)))
allyears <- do.call(rbind, tmp)

# see what that looks like:
head(allyears)
```

Now, we can plot it like before, but `facet_wrap` on date:
```{r sst-facets, fig.height=11, fig.width=8, out.height="1150px", out.width="750px"}
ggplot(data = allyears, aes(x = long, y = lat, fill = degF)) + 
  geom_raster(interpolate = TRUE) +
  geom_polygon(data = shore, aes(x = long, y = lat, group = group), color = "black", fill = "grey80") +
  scale_fill_gradientn(colours = rev(rainbow(7)), na.value = NA) +
  theme_bw() +
  coord_fixed(1.3) + 
  facet_wrap(~ date, nrow = 3)
```

That takes a minute or so to plot on my old laptop---not super zippy, but it is
so easy to code up and it looks pretty good!  It is also clear that my buddies
have it right: the water really is quite warm this year.

I'm outta here! I've gotta pack my mask and snorkel up...



## Failures with ggmap... 

Now, what if we want a ggmap base to it?  The following causes a complete failure. R aborts....

```{r, eval=FALSE}
library(ggmap)

bb <- make_bbox(lon, lats)
base <- get_map(bb, maptype = "terrain-background")


ggmap(base) + 
  geom_raster(data = msst, aes(x = long, y = lat, fill = value, alpha = not_missing), interpolate = TRUE) +
  scale_fill_gradientn(colours = rev(rainbow(7))) + 
  scale_alpha_discrete(range = c(0,1))

```





So, what if we transform the points to mercator first?
```{r, eval=FALSE}
library(ggmap)
library(rgdal)
library(sp)

bb <- make_bbox(lon, lats)
base <- get_map(bb, maptype = "terrain-background")

# let's try to project it to the mercator:
# make a SpatialPointsDataFrame of it.
sst_spdf <- SpatialPointsDataFrame(msst[,1:2], data = msst[, -(1:2)], proj4string = CRS("+proj=longlat +datum=WGS84"))

sst2 <- spTransform(sst_spdf, CRS("+proj=merc"))


ggmap(base) + 
  geom_raster(data = as.data.frame(sst2), aes(x = long, y = lat, fill = value, alpha = not_missing), interpolate = TRUE) +
  scale_fill_gradientn(colours = rev(rainbow(7))) + 
  scale_alpha_discrete(range = c(0,1))

```
Nope! That totally bombs too!

