---
title: Comparing Map Resolutions
author: Eric C. Anderson
date: '2014-12-17'
slug: comparing-map-resolutions
categories: []
tags: []
---

---
title: "Comparing Map Resolutions"
output: html_document
author: Eric C. Anderson
keep-md: true
clean: false

---


I am curious about how the resolutions of shorelines differ between different map sources.

Roy M just pointed me to NOAA's [GSHHS](http://www.ngdc.noaa.gov/mgg/shorelines/gshhs.html) shoreline maps (they also
have boundaries and rivers).  I am curious how that compares to the `worldHires` map available in `mapdata`. 
I downloaded the 113 Mb "bin" version that was at [http://www.ngdc.noaa.gov/mgg/shorelines/data/gshhg/latest/gshhg-bin-2.3.3.zip](http://www.ngdc.noaa.gov/mgg/shorelines/data/gshhg/latest/gshhg-bin-2.3.3.zip).  I put the resulting
directory inside a directory called `Maps` in my home directory.

Load all the libraries we will use:
```{r, message=FALSE}
library(mapdata)
library(maptools)
library(ggplot2)
library(dplyr) 
```


## Looking at worldHiRes

Let's have a look at SF bay, first at a California scale:
```{r sf-md-1}
w <- map_data("worldHires", ylim = c(35,40), xlim = c(-125,-120))
ggplot() + 
  geom_polygon(data = w, aes(x = long, y = lat, group = group), fill = "grey80") + 
  coord_fixed(1.3, xlim = c(-125, -120), ylim = c(35, 40)) +   # wrote this before coord_quickmap()...
  theme_bw()
```

And then at a Bay scale:
```{r sf-md-2}
ggplot() + 
  geom_polygon(data = w, aes(x = long, y = lat, group = group), fill = "grey80") + 
  coord_fixed(1.3, xlim = c(-123, -122), ylim = c(37.5, 38.5)) + 
  theme_bw()
```

That is pretty chunky.


## And compare to GSHHS

First at the California scale.
```{r, message=FALSE, error=FALSE}
if (!rgeosStatus()) gpclibPermit()
gshhs.f.b <- "/Users/eriq/Maps/gshhg-bin-2.3.3/gshhs_f.b"
sf1 <- getRgshhsMap(gshhs.f.b, xlim = c(-125, -120), ylim = c(35, 40)) %>%
  fortify()

ggplot() + 
  geom_polygon(data = sf1, aes(x = long, y = lat, group = group), fill = "grey80") + 
  coord_fixed(1.3, xlim = c(-125, -120), ylim = c(35, 40)) + 
  theme_bw()

```

Then at the Bay scale:
```{r}
ggplot() + 
  geom_polygon(data = sf1, aes(x = long, y = lat, group = group), fill = "grey80") + 
  coord_fixed(1.3, xlim = c(-123, -122), ylim = c(37.5, 38.5)) + 
  theme_bw()
```

OK.  So the GSHHS at full resolution is clearly higher quality.


## Another important thing to note

What is even more intriguing about this is that it appears that ``ggplot2`'s `get_map` function is not very good
at grabbing just the part of the `worldHires` map that is requested.  When you ask for the SF bay region it returns
pretty much all of the USA, perhaps because that is all one polygon. On the other hand, when using `getRgshhsMap` it
it appears that the function grabs just the area of interest and then appropriately closes the polygons.   As a
consequence, the GSHHS maps seems to plot a little faster using `ggplot` than the `mapdata` map, and it takes
up much less space in memory:
```{r}
# how big is the GSHHS map data?
format(object.size(sf1), units = "Mb")

# what about worldHiRes?
format(object.size(w), units = "Mb")
```


## Let us add some rivers
I am curious to know at what scale the rivers are mapped to.
```{r}
wdb_rivers_f.b <- "/Users/eriq/Maps/gshhg-bin-2.3.3/wdb_rivers_f.b"
rivers <- getRgshhsMap(wdb_rivers_f.b, xlim = c(-125, -120), ylim = c(35, 40)) 
```

oops! outta time.  Maybe come back to this later...

## Session Info
```{r}
sessionInfo()
```
